% !TeX root = ../../book.tex
\section{关系}
\secbegin{secRelations}
\index{relation|(}

数学中许多有趣的结论或概念都是通过观察集合间元素的交互，或单个集合中元素间的交互产生的。我们可以用\textit{关系}这一概念让`交互'的概念变得精确。

\begin{definition}
\index{relation}
\index{homogeneous relation}
\index{domain!of a relation}
\index{codomain!of a relation}
设 $X$ 和 $Y$ 为集合。$X$ 到 $Y$ 的（\textbf{二元}）\textbf{关系}为逻辑表达式 $R(x,y)$，其中两个自由变量为 $x \in X$ 和 $y \in Y$。我们称 $X$ 为 $R$ 的\textbf{定义域}，$Y$ 为 $R$ 的\textbf{值域}。

如果关系 $R$ 具有相同的定义域和值域，则该关系是 \textbf{同构的}，在这种情况下，我们说 $R$ 是 $X$ \textbf{上}的关系。

给定 $x \in X$ 和 $y \in Y$，如果 $R(x,y)$ 为真，则我们说`$x$ 通过 $R$ 与 $y$ \textbf{关联}'，写做 $x \mathrel{R} y$ \inlatexnb{x \textbackslash{}mathrel\{R\} y}\lindexmmc{mathrel}{$\mathrel{R}, \dots$}。
\end{definition}

\begin{example}
\label{exRelationExamples}
我们已经见过很多关系的例子。
\begin{itemize}
\item 整除性（`$x$ 除 $y$'） 是 $\mathbb{Z}$ 上的关系。
\item 不等关系 $\le$ 是 $\mathbb{R}$ 上的关系。
\item 对于任意集合 $X$，相等 $=$ 是 $X$ 上的关系。
\item 逻辑等价 $\equiv$ 是所有逻辑公式集合上的关系。
\item 对于任意集合 $X$，子集关系 $\subseteq$ 是 $\mathcal{P}(X)$ 上的关系。
\end{itemize}
这些关系都是同构的，但并非所有关系都如此：
\begin{itemize}
\item 对于任意集合 $X$, 元素关系 $\in$ 是从 $X$ 到 $\mathcal{P}(X)$ 的关系。
\item 每个函数 $f : X \to Y$ 通过将 $x \mathrel{R_f} y$ 定义为 $f(x) = y$ 引入从 $X$ 到 $Y$ 的关系 $R_f$。
\end{itemize}
\end{example}

\begin{exercise}
再举三个关系的例子，并且不要所都是同构的。
\end{exercise}

与集合和函数一样，我们必须判定合适可以说两个关系是相等的。例如，考虑 $\mathbb{R}$ 上的关系 $R$，对于 $a,b \in \mathbb{R}$ 令 $a \mathrel{R} b$ 表示 $\exists x \in \mathbb{R},~ a+x^2=b$。刚好 $a \mathbin{R} b$ 当且仅当 $a \le b$ --- 我们将在 \Cref{exLeqOnRFromAddingSquare} 中证明这一点。那么 $R$ 与 $\le$ 相等吗？一方面，你可能会说`是'，因为 $\le$ 和 $R$ 关联相同的实数对。另一方面，你可能会说`不'，因为 $\le$ 和 $R$ 关联相同的实数对这一事实并不是显而易见的，需要证明。事实上，如果我们用 $\mathbb{Q}$ 替换 $\mathbb{R}$，那么 $\le$ 和 $R$ 将\textit{不}关联相同的元素对，因为例如 $0 \le 2$ 但不存在满足 $0+x^2=2$ 的有理数 $x$。

然而与集合和函数一样，关系满足相等的\textit{外延}概念：就像当两个集合具有相同的元素时相等（\Cref{axSetEquality}），当两个函数具有相同的值时相等（\Cref {axFunctionExtensionality}）一样，当两个关系涉及相同的元素对时，我们认为两个关系是相等的（\Cref{axRelationEquality}）。

\begin{axiom}[关系外延性]
\label{axRelationEquality}
设 $R$ 和 $S$ 为关系。则 $R = S$ 当且仅当 $R$ 和 $S$ 具有相同的定义域 $X$ 和值域 $Y$，且
\[ \forall x \in X,\, \forall y \in Y,\, (x \mathrel{R} y \Leftrightarrow x \mathrel{S} y) \]
也就是说，具有相同定义域和值域的两个关系在关联相同元素对时严格相等。
\end{axiom}

\begin{example}
\label{exLeqOnRFromAddingSquare}
回想一下我们上面定义的 $\mathbb{R}$ 上的关系 $R$，对于 $a,b \in \mathbb{R}$ 令 $a \mathrel{R} b$ 当且仅当 $a+x^2=b$，其中 $x \in \mathbb{R}$。要证明 $R = {\le}$，注意到 $a + (b-a) = b$，并且 $b-a$ 为实数的平方当且仅当 $b-a \ge 0$，即当且仅当 $a \le b$。
\end{example}

\begin{exercise}
设 $R$ 和 $S$ 为 $\mathbb{R}$ 上的关系，对于 $a,b \in \mathbb{R}$ 令
\[ a \mathrel{R} b ~ \Leftrightarrow ~ b-a \in \mathbb{Q} \quad \text{且} \quad a \mathrel{S} b ~ \Leftrightarrow ~ \exists n \in \mathbb{Z},~ (n \ne 0) \wedge n(b-a) \in \mathbb{Z} \]
证明 $R=S$。
\end{exercise}

The true reason why \Cref{axRelationEquality} is powerful is that it allows us to reason about relations entirely set theoretically by working with their \textit{graphs}---the sets of pairs of elements that they relate---rather than with the particular formulae defining the relation.

\begin{definition}
\index{graph!of a relation}
\label{defRelationGraph}
Let $R$ be a relation from a set $X$ to a set $Y$. The \textbf{graph} of $R$ is the set $\mathrm{Gr}(R)$\nindex{graphRelation}{$\mathrm{Gr}(R)$}{graph of a relation} \inlatex{mathrm\{Gr\}\{R\}}\lindexmmc{mathrm}{$\mathrm{Aa}, \mathrm{Bb}, \dots$} of pairs $(x,y) \in X \times Y$ for which $x \mathrel{R} y$. That is
\[ \mathrm{Gr}(R) = \{ (x,y) \in X \times Y \mid x \mathrel{R} y \} \subseteq X \times Y \]
\end{definition}

\begin{example}
The graph of the relation $\le$ on $[3]$ is
\[ \{ (1,1), (1,2), (1,3), (2,2), (2,3), (3,3) \} \]
Likewise, the graph of the relation $\le$ viewed as a relation from $[2]$ to $[4]$ is
\[ \{ (1,1), (1,2), (1,3), (1,4), (2,2), (2,3), (2,4) \} \]
This demonstrates that the graph of a relation is sensitive to the domain (and codomain) of the relation.
\end{example}

\begin{example}
\label{exCircleRelation}
Consider the relation $C$ from $\mathbb{R}$ to $\mathbb{R}$ defined by $x \mathrel{C} y \Leftrightarrow x^2+y^2=1$. Then
\[ \mathrm{Gr}(C) = \{ (x,y) \in \mathbb{R} \times \mathbb{R} \mid x^2+y^2=1 \} \]
Plotting $\mathrm{Gr}(C)$ on a standard pair of axes yields a circle with radius $1$ centred at the point $(0,0)$, shown below with a unit grid.
\begin{center}
\begin{tikzpicture}
\foreach \i in {-1,1} {
    \draw[lightgray] (-3.3,{2*\i}) -- (3.3,{2*\i}) ;
    \draw[lightgray] ({2*\i},-3.3) -- ({2*\i},3.3) ;
}
\draw[-latex] (-3.3,0) -- (3.3,0) node[right] {$x$} ;
\draw[-latex] (0,-3.3) -- (0,3.3) node[right] {$y$} ;
\draw[blue] (0,0) circle[radius=2] ;
\end{tikzpicture}
\end{center}
Note that $\mathrm{Gr}(C)$ is \textit{not} the graph of a function $f : [0,1] \to \mathbb{R}$, since for example both $(0,1)$ and $(0,{-1})$ are elements of $\mathrm{Gr}(C)$, the value $f(0)$ would not be uniquely defined.
\end{example}

\begin{exercise}
Let $R$ be the relation on $\mathbb{Z}$ defined for $x,y \in \mathbb{Z}$ by letting $x \mathrel{R} y$ if and only if $x^2=y^2$. Describe its graph $\mathrm{Gr}(R) \subseteq \mathbb{Z} \times \mathbb{Z}$.
\end{exercise}

\begin{exercise}
Let $f : X \to Y$ be a function, and define the relation $R_f$ from $X$ to $Y$ as in \Cref{exRelationExamples}. Prove that $\mathrm{Gr}(R_f) = \mathrm{Gr}(f)$---that is, the graph of the \textit{relation} $R_f$ is equal to the graph of the \textit{function} $f$.
\end{exercise}

\begin{definition}
\label{defEmptyRelation}
\label{defDiscreteRelation}
\index{empty relation}
\index{discrete relation}
\index{relation!empty}
\index{relation!discrete}
The \textbf{discrete relation} from a set $X$ to a set $Y$ is the relation $D_{X,Y}$ defined by letting $x \mathrel{D_{X,Y}} y$ be true for all $x \in X, y \in Y$.

The \textbf{empty relation} from a set $X$ to a set $Y$ is the relation $\varnothing_{X,Y}$\nindex{OXY}{$\varnothing_{X,Y}$}{empty relation} \inlatex{varnothing} defined by letting $x \mathrel{\varnothing_{X,Y}} y$ be false for all $x \in X, y \in Y$.
\end{definition}

\begin{exercise}
\label{exGraphOfEmptyAndDiscreteRelations}
Let $X$ and $Y$ be sets. Describe the graphs $\mathrm{Gr}(D_{X,Y})$ and $\mathrm{Gr}(\varnothing_{X,Y})$.
\end{exercise}

It turns out that, for fixed sets $X$ and $Y$, relations from $X$ to $Y$ correspond with subsets of $X \times Y$---see \Cref{thmRelationsAsSubsets} below. This fact is so convenient that many (if not most) authors actually define `relation from $X$ to $Y$' to mean `subset of $X \times Y$'.

\begin{theorem}
\label{thmRelationsAsSubsets}
Let $X$ and $Y$ be sets. Every subset $G \subseteq X \times Y$ is the graph of a unique relation $R$ from $X$ to $Y$.
\end{theorem}
\begin{cproof}
Fix $G \subseteq X \times Y$. Define a relation $R$ by
\[ \forall x \in X,\, \forall y \in Y,\, x \mathbin{R} y \Leftrightarrow (x,y) \in G \]
Then certainly $G = \mathrm{Gr}(R)$, since for all $x \in X$ and $y \in Y$ we have
\[ (x,y) \in G \quad \Leftrightarrow \quad x \mathrel{R} y \quad \Leftrightarrow \quad (x,y) \in \mathrm{Gr}(R) \]

Moreover, if $S$ is a relation from $X$ to $Y$ such that $G=\mathrm{Gr}(S)$, then, for all $x \in X$ and $y \in Y$ 
\[ x \mathrel{S} y \Leftrightarrow (x,y) \in \mathrm{Gr}(S) \Leftrightarrow (x,y) \in G \Leftrightarrow x \mathrel{R} y \]
so $S=R$.

Hence there is exactly one relation from $X$ to $Y$ whose graph is $G$.
\end{cproof}

\Cref{thmRelationsAsSubsets} suggests that, for the purposes of defining relations and proving that relations are equal, we may work entirely set theoretically with the graphs of the relations.

\begin{strategy}[Relations as graphs]
\label{strRelationsAndGraphs}
In order to specify a relation $R$, it suffices to specify its domain $X$, its codomain $Y$, and its graph $\mathrm{Gr}(R) \subseteq X \times Y$. Furthermore, in order to prove that two relations $R$ and $S$ are equal, it suffices to prove that they have the same domain and codomain, and that their graphs are equal.
\end{strategy}

\begin{example}
Consider the set $G = \{ (2m+i, 2n+i) \mid m,n \in \mathbb{Z},~ i \in \{ 0,1 \} \}$. Since $G \subseteq \mathbb{Z} \times \mathbb{Z}$, it is the graph of a (unique) relation $R$ on $\mathbb{Z}$, which is necessarily defined for $a,b \in \mathbb{Z}$ by letting $a \mathrel{R} b$ if and only if there are integers $m$ and $n$, and $i \in \{0,1\}$, such that $a=2m+i$ and $b=2n+i$. But this says precisely that $a$ and $b$ both leave the same remainder (namely $i$) when divided by $2$, so that $R$ can be described by saying that, for all $a,b \in \mathbb{Z}$, we have $a \mathrel{R} b$ if and only if $a$ and $b$ are both even or both odd.
\end{example}

\begin{definition}
\label{defDiagonalSubset}
\index{diagonal subset}
\index{subset!diagonal}
Let $X$ be a set. The \textbf{diagonal subset} of $X \times X$ is the set $\Delta_X$ \inlatex{Delta\_X} defined by $\Delta_X = \{ (x,x) \mid x \in X \}$.
\end{definition}

To see why $\Delta_X$ is called the `diagonal' subset, try plotting $\Delta_{\mathbb{R}} \subseteq \mathbb{R} \times \mathbb{R}$ on a standard pair of axes (like in \Cref{exCircleRelation}).

\begin{exercise}
Let $X$ be a set. Prove that $\Delta_X = \mathrm{Gr}(=)$.
\end{exercise}

\subsection*{齐次关系的性质}

Most of the relations of interest to us in this book are homogeneous---that is, relations \textit{on} a set. In fact, they broadly fall into one of two categories: \textit{equivalence relations}, which are relations that `behave like $=$'; and \textit{order relations}, which are relations that `behave like $\le$'. We will study equivalence relations in \Cref{secEquivalenceRelationsPartitions} and order relations in \Cref{secOrderRelations}, but examples of such relations pop up throughout the book. (In fact, we have already seen several!)

Our task for the rest of this section is to isolate the properties that a relation must satisfy in order to be classified as an equivalence relation or an order relation.

To aid with intuition, we will illustrate these properties with diagrams: given a relation $R$, the fact that $a \mathrel{R} b$ will be represented diagramatically as follows:
\begin{center}
\begin{tikzpicture}
\draw (0,0) \relnode{a}{$a$} ;
\draw (3,0) \relnode{b}{$b$} ;
\draw[-latex] (a) -- (b) ;
\end{tikzpicture}
\end{center}

A \textit{reflexive} relation is one that relates every element of its domain to itself.

\begin{definition}
\label{defReflexiveRelation}
\index{reflexive relation}
\index{relation!reflexive}
A relation $R$ on a set $X$ is \textbf{reflexive} if $a \mathrel{R} a$ for all $a \in X$.
\end{definition}

\begin{center}
\begin{tikzpicture}
\draw (0,0) \relnode{a}{$a$} ;
\draw[-latex] (a) to[loop] (a) ;
\end{tikzpicture}
\end{center}

\begin{example}
\label{exEqualityIsReflexive}
Given any set $X$, the equality relation $=$ on $X$ is reflexive, since $a = a$ for all $a \in X$.
\end{example}

\begin{example}
\label{exDifferenceOfRealsInQIsReflexive}
%% BEGIN EXTRACT %%
Let $R$ be the relation on $\mathbb{R}$ defined for $a,b \in \mathbb{R}$ by $a \mathrel{R} b$ if and only if $b-a \in \mathbb{Q}$. Then $R$ is reflexive, since for all $a \in \mathbb{R}$, we have $a-a = 0 \in \mathbb{Q}$, so that $a \mathrel{R} a$.
%% END EXTRACT %%
\end{example}

\begin{exercise}
\label{exSubsetIsReflexive}
Let $X$ be a set. Prove that $\subseteq$ is a reflexive relation on $\mathcal{P}(X)$, but $\subsetneqq$ is not.
\end{exercise}

\begin{exercise}
\label{exDivisibilityIsReflexive}
Prove that the relation `$x$ divides $y$' on $\mathbb{Z}$ is reflexive.
\end{exercise}

The next exercise demonstrates that when determining if a relation is reflexive, we must be careful to specify its domain.

\begin{exercise}
\label{exReflexivitySensitiveToDomainOfRelation}
Let $G = \{ (1,1), (2,2), (3,3) \}$. Let $R$ be the relation on $[3]$ whose graph is $G$, and let $S$ be the relation on $[4]$ whose graph is $G$. Prove that $R$ is reflexive, but $S$ is not.
\end{exercise}

Symmetric relations are those for which the \textit{direction} of the relation doesn't matter: two elements are either each related to the other, or not related at all.

\begin{definition}
\label{defSymmetricRelation}
\index{symmetric relation}
\index{relation!symmetric}
A relation $R$ on a set $X$ is \textbf{symmetric} if, for all $a,b \in X$, if $a \mathrel{R} b$, then $b \mathrel{R} a$.
\end{definition}

\begin{center}
\begin{tikzpicture}
\draw (0,0) \relnode{a}{$a$} ;
\draw (3,0) \relnode{b}{$b$} ;
\draw[-latex] (a) to[bend left] (b) ;
\draw[-latex, dashed] (b) to[bend left] (a) ;
\end{tikzpicture}
\end{center}

\begin{example}
\label{exEqualityIsSymmetric}
Given any set $X$, the equality relation $=$ on $X$ is symmetric, since for all $a,b \in X$, if $a=b$, then $b=a$.
\end{example}

\begin{example}
\label{exDifferenceOfRealsInQIsSymmetric}
Let $R$ be the relation on $\mathbb{R}$ defined for $a,b \in \mathbb{R}$ by $a \mathrel{R} b$ if and only if $b-a \in \mathbb{Q}$. Then $R$ is symmetric.

To see this, let $a,b \in \mathbb{R}$ and assume that $a \mathrel{R} b$. Then $b-a \in \mathbb{Q}$, so that $b-a = \dfrac{p}{q}$ for some $p,q \in \mathbb{Z}$ with $q \ne 0$. But then
\[ a-b = -(b-a) = \dfrac{-p}{q} \]
so that $a-b \in \mathbb{Q}$. Hence $b \mathrel{R} a$, as required.
\end{example}

\begin{exercise}
Find all subsets $U \subseteq \mathbb{Z}$ such that the relation `$x$ divides $y$' on $U$ is symmetric.
\end{exercise}

We showed in \Cref{exReflexivitySensitiveToDomainOfRelation} that reflexivity of a relation is sensitive to its domain. The next exercise demonstrates that symmetry is \textit{not} sensitive to the domain---that is, it is an \textit{intrinsic} property of the relation.

\begin{exercise}
\label{exSymmetryNotSensitiveToDomainOfRelation}
Let $R$ and $S$ be relations such that $\mathrm{Gr}(R) = \mathrm{Gr}(S)$. Note that the domain of $R$ might be different from the domain of $S$. Prove that $R$ is symmetric if and only if $S$ is symmetric.
\end{exercise}

A condition related to symmetry, but in a sense opposite to it, is \textit{antisymmetry}. It says that the only way that two elements of a set can each be related to the other is if they are equal.

\begin{definition}
\index{antisymmetric relation}
\index{relation!antisymmetric}
Let $X$ be a set. A relation $R$ on $X$ is \textbf{antisymmetric} if, for all $a,b \in X$, if $a \mathrel{R} b$ and $b \mathrel{R} a$, then $a=b$.
\end{definition}

\begin{center}
\begin{tikzpicture}
\draw (0,0) \relnode{a}{$a$} ;
\draw (3,0) \relnode{b}{$b$} ;
\draw[-latex] (a) to[bend left] (b) ;
\draw[-latex] (b) to[bend left] (a) ;
\draw (4.5,0) node {$\Rightarrow$} ;
\draw (6,0) node {$a=b$} ;
\end{tikzpicture}
\end{center}

A word of warning here is that `antisymmetric' does not mean the same thing as `not symmetric'---indeed, we we will see, equality is both symmetric and antisymmetric, and many relations are neither symmetric nor antisymmetric. [Even more confusingly, there is a notion of \textit{asymmetric relation}, which also does not mean `not symmetric'.]

\begin{example}
Given any set $X$, the equality relation $=$ on $X$ is antisymmetric, since for all $a,b \in X$, if $a=b$ and $b=a$, then $a=b$.
\end{example}

\begin{example}
The order relation $\le$ on $\mathbb{R}$ is antisymmetric, since for all $a,b \in \mathbb{R}$, if $a \le b$ and $b \le a$, then $a=b$.
\end{example}

\begin{exercise}
\label{exDivisibilityIsOrIsNotAntisymmetric}
Prove that the relation `$x$ divides $y$' on $\mathbb{N}$ is antisymmetric, but not on $\mathbb{Z}$.
\end{exercise}

\begin{exercise}
Let $X$ be a set. Prove that $\subseteq$ is an antisymmetric relation on $\mathcal{P}(X)$.
\end{exercise}

\begin{exercise}
Let $X$ be a set and let $R$ be a relation on $X$. Prove that $R$ is both symmetric and antisymmetric if and only if $\mathrm{Gr}(R) \subseteq \Delta_X$, where $\Delta_X$ is the diagonal subset of $X \times X$ (see \Cref{defDiagonalSubset}). Deduce that the only reflexive, symmetric and antisymmetric relation on a set $X$ is the equality relation on $X$.
\end{exercise}

The last property we will study in some detail is \textit{transitivity}. Transitive relations are those for which we can skip over intermediate related elements---for example, we can deduce $0 < 3$ from the facts that $0 < 1$ and $1 < 2$ and $2 < 3$.

\begin{definition}
\label{defTransitiveRelation}
\index{relation!transitive}
\index{transitive relation}
A relation $R$ on a set $X$ is \textbf{transitive} if, for all $a,b,c \in X$, if $a \mathrel{R} b$ and $b \mathrel{R} c$, then $a \mathrel{R} c$.
\end{definition}

\begin{center}
\begin{tikzpicture}
\draw (0,0) \relnode{a}{$a$} ;
\draw (3,0) \relnode{b}{$b$} ;
\draw (6,0) \relnode{c}{$c$} ;
\draw[-latex] (a) to (b) ;
\draw[-latex] (b) to (c) ;
\draw[-latex, dashed] (a) to[bend left] (c) ;
\end{tikzpicture}
\end{center}

\begin{example}
\label{exEqualityIsTransitive}
Given any set $X$, the equality relation $=$ on $X$ is transitive since, for all $a,b,c \in X$, if $a=b$ and $b=c$, then $a=c$.
\end{example}

\begin{example}
\label{exDifferenceOfRealsInQIsTransitive}
Let $R$ be the relation on $\mathbb{R}$ defined for $a,b \in \mathbb{R}$ by $a \mathrel{R} b$ if and only if $b-a \in \mathbb{Q}$. Then $R$ is transitive.

To see this, let $a,b,c \in \mathbb{R}$ and assume that $a \mathrel{R} b$ and $b \mathrel{R} c$. Then $b-a \in \mathbb{Q}$ and $c-b \in \mathbb{Q}$, so there exist $p,q,r,s \in \mathbb{Z}$ with $q,s \ne 0$ such that
\[ b-a = \dfrac{p}{q} \quad \text{and} \quad c-b = \dfrac{r}{s} \]
It follows that 
\[ c-a = (c-b) + (b-a) = \dfrac{p}{q} + \dfrac{r}{s} = \dfrac{ps+qr}{qs} \]
so that $c-a \in \mathbb{Q}$. Hence $a \mathrel{R} c$, as required.
\end{example}

\begin{exercise}
Let $X$ be a set. Prove that $\subseteq$ is a transitive relation on $\mathcal{P}(X)$.
\end{exercise}

\begin{exercise}
\label{exDivisibilityIsTransitive}
Prove that the relation `$x$ divides $y$' on $\mathbb{Z}$ is transitive.
\end{exercise}

Like symmetry, transitive is an intrinsic property of relations---that is, transitivity is not sensitive to the domain of the relation---as the next exercise demonstrates.

\begin{exercise}
\label{exTransitivityNotSensitiveToDomainOfRelation}
Let $R$ and $S$ be relations such that $\mathrm{Gr}(R) = \mathrm{Gr}(S)$. Note that the domain of $R$ might be different from the domain of $S$. Prove that $R$ is transitive if and only if $S$ is transitive.
\end{exercise}

A fundamental property of transitive relations is that we can prove two elements $a$ and $b$ are related by finding a chain of related elements starting at $a$ and finishing at $b$. This is the content of the following proposition.

\begin{proposition}
\label{propTransitiveChain}
Let $R$ be a relation on a set $X$. Then $R$ is transitive if and only if, for any finite sequence $x_0, x_1, \dots, x_n$ of elements of $X$ such that $x_{i-1} \mathrel{R} x_i$ for all $i \in [n]$, we have $x_0 \mathrel{R} x_n$.
\end{proposition}
\begin{cproof}
For the sake of abbreviation, let $p(n)$ be the assertion that, for any $n \ge 1$ and any sequence $x_0, x_1, \dots, x_n$ of elements of $X$ such that $x_{i-1} \mathrel{R} x_i$ for all $i \in [n]$, we have $x_0 \mathrel{R} x_n$.

We prove the two directions of the proposition separately.
\begin{itemize}
\item ($\Rightarrow$) Suppose $R$ is transitive. For $n \ge 1$. We prove $p(n)$ is true for all $n \ge 1$ by induction.
\begin{itemize}
\item (\textbf{Base case}) When $n=1$ this is immediate, since we assume that $x_0 \mathrel{R} x_1$.
\item (\textbf{Induction step}) Fix $n \ge 1$ and suppose $p(n)$ is true. Let $x_0, \dots, x_n, x_{n+1}$ is a sequence such that $x_{i-1} \mathrel{R} x_i$ for all $i \in [n+1]$. We need to prove that $x_0 \mathrel{R} x_{n+1}$.

By the induction hypothesis we know that $x_0 \mathrel{R} x_n$. By definition of the sequence we have $x_n \mathrel{R} x_{n+1}$. By transitivity, we have $x_0 \mathrel{R} x_{n+1}$.
\end{itemize}
So by induction, we have proved the $\Rightarrow$ direction.
\item ($\Leftarrow$)
%% BEGIN EXTRACT (xtrSubstitutionExample) %%
Suppose $p(n)$ is true for all $n \ge 1$. Then in particular $p(2)$ is true, which is precisely the assertion that $R$ is transitive.
%% END EXTRACT %%
\end{itemize}
So we're done.
\end{cproof}

That is, \Cref{propTransitiveChain} states that for a transitive relation $R$ on a set $X$, if $x_0, x_1, \dots, x_n \in X$, then
\[ x_0 \mathrel{R} x_1 \mathrel{R} \cdots \mathrel{R} x_n \quad \Rightarrow \quad x_0 \mathrel{R} x_n \]
where `$x_0 \mathrel{R} x_1 \mathrel{R} \cdots  \mathrel{R} x_n$' abbreviates the assertion that $x_i \mathrel{R} x_{i+1}$ for each $i < n$.

\index{relation|)}